Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-02 paper2.qmd
section:6; process:write

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines Causal and Bayesian inference methods. In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear in the rest of the document clarifies these concepts. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers rely on estimates from the BTL model, or its transformations, to approximate this estimand.

<!-- 4. But why identification is so important? -->
However, @sec-theory-issues presents compelling evidence that Thurstone's Case V, and by extension the BTL model, suffers from several statistical and measurement limitations. These limitations hinder the model’s ability to identify various estimands relevant to CJ inquiries, including the one described in the example. Identification is crucial because it is a necessary condition for ensuring consistent estimators. *Consistency* refers to the property of an estimator whose estimates converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010]. Without identification, consistency cannot be achieved, even with "infinite" and error-free data. Thus, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023].

@@

<!-- 6. What the section will do? -->
Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to Causal inference. Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
$$
I = \begin{bmatrix}
1 \\
\vdots \\
i \\
\vdots \\
h \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
i & a \\
\vdots & \vdots \\
h & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A}
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K}
\end{bmatrix}
$$ {#eq-mat11}

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $i$ or $h$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $i, h$ with stimuli $a, b$. As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a (repeated) judgment index $k$. Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
$$
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
i & a & h & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
\end{bmatrix}
$$ {#eq-mat12}

Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix. This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population. Thus, we define the matrix as in Equation ([-@eq-mat12]).

It is easier to visualize the structure of the previously defined vectors and matrices by considering an example. Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the form as in Equation ([-@eq-mat13]).
$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
$$ {#eq-mat13}

@@

We can illustrate the structure of these vectors more clearly with an example. Suppose researchers exclude the second student, the second text from each student, and the third judge from the setup shown in Equation ([-@eq-mat13]). Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the resulting vectors would have the following structure:
$$
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
1 \\
1 \\
1 \\
1 \\
0 \\
0 \\
0 
\end{bmatrix}
$$ {#eq-mat24}

@@

Finally, we define the *sample mechanism* $S$ in Equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. Each element $s_{(i,a,h,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in Equation ([-@eq-mat26]). Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
$$
S = \begin{bmatrix}
s_{(1,1,1,2,1,1)} \\
\vdots \\
s_{(1,1,1,2,1,n_{K})} \\
\vdots \\
s_{(i,a,h,b,j,k)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
$$ {#eq-mat25}

@@

@@

<!-- 4. General summary of what we have done. -->
Thus, to address these limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods. The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.

@@

<!-- 5. Are there any additional advantages? -->
Beyond these potential benefits, the approach offers two additional advantages. First, it clarifies the roles and interactions of all actors and processes involved in CJ experiments. Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data. Together, these advantages establish a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry.

@@

Five research avenues, both current and new, deserve particular attention. The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them. Notably, although it may seem like a natural next step to address these questions through formal derivations, following the tradition of classical statistical models like linear regression, we contend that the complexity of the CJ system makes this approach impractical from the outset. Instead, simulation-based methods, such as power analysis, offer a more feasible and flexible alternative, enabling researchers to investigate these questions without depending on cumbersome mathematical proofs. Nonetheless, developing formal mathematical proofs remains an essential goal for future work.

<!-- due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences. -->

@@

<!-- 1. Challenges for the framework: background knowledge and research context -->
Despite its promise, this study presents several challenges that researchers must navigate --challenges common to many quantitative studies. These challenges fall into two main categories. The first concerns the background knowledge researchers need to apply the proposed framework effectively. The second involves the increased attention researchers must now devote to their specific CJ research context.

<!-- 2. Challenge 1: background knowledge -->
In terms of background knowledge, researchers face two challenges: mastering Causal inference and developing proficiency in Bayesian inference methods. One example illustrating the need to master Causal inference is our assumption that predictor variables are "relevant" to the research context--interpreting relevance as the inclusion of these variables in a "sufficient adjustment set" [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. However, we do not fully explore what this entails or its implications for model specification, estimation, and inference. As a result, researchers must engage deeply with these complex ideas--and with others, such as how SCMs encode functional and probabilistic information--to apply the framework rigorously. To assist this effort, we provide key references throughout the study to guide motivated readers toward a deeper understanding of these foundational concepts[^9].

@@

@@

Bayesian inference presents a second challenge. Researchers must acquire sufficient expertise to translate context-specific conceptual models into bespoke statistical models that meaningfully address their research questions. Although such translations are not unique to Bayesian approaches, we argue that Bayesian methods offer a more accessible route, as they reduce the need for specialized knowledge in optimization theory and techniques--often required by frequentist methods--thereby lowering the technical barriers for many applied researchers.

<!-- 3. Challenge 2: research context -->
In terms of the attention researchers must now devote to their specific CJ research context, researchers face two challenges: ensuring the conceptual model is a *faithful* representation for the CJ system, and verifying whether this model can accurately estimate the intended estimand from real data.

<!-- See fine point 6.1 and 6.2 [@Hernan_et_al_2025, pp. 79-81] -->

<!-- we are confident that these causal structure is faithful to a design -->

<!-- Ideally, we would like to assess this property through an *identification analysis* [@Schuessler_et_al_2023], where researchers assess how accurate a model computes an estimand based solely on its (causal) assumptions, regardless of random variability. As one can imagine these types of analysis requires formal proofs, nevertheless, we argue that simulation-based methods can provide preliminary assessments akin to an identification analysis. -->

<!-- In this regard, we argue that researchers face the challenge that need a thorough knowledge about the CJ system, particularly regarding experimental design, the assumptions underlying the data analysis, and appropriate model specification. This is neccesary because the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions. For instance, the selection of relevant variables largely depends on the research context, and thus you really have to think through which variables you have to put in in your analysis. -->

<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable. Nevertheless, as @Kohler_et_al_2019 [pp. 150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data. -->

<!-- Nevertheless, although in this paper we have outlined the causal process behind the CJ system, our framework is not designed to work every CJ design or data, is just a starting point for researchers to think thoroughly about the causal structure and statistical model that applies for their research context. That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly. -->

<!-- Fifth and final, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate. In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest? However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces. Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement. -->

@@

<!-- Naturally, the validity of causal claims now hinges on the assumption that these variables serve as a sufficient adjustment set, a property that cannot be guaranteed for any variable. Nevertheless, as @Kohler_et_al_2019 [pp. 150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions, and these requirements hold true for both observational and experimental data. -->

@@

@@

<!-- 1. How the SCMS and DAGs encode probabilistic information? -->
Beyond their graphical capabilities, SCMs and DAGs can encode the probabilistic information embedded within a causal structure. They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges assumption. The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020, pp. 20]. Meanwhile, the *minimality assumption* encodes probabilistic dependencies among variables by stating that every pair of adjacent nodes exhibits a dependency [@Neal_2020, pp. 21]. Finally, the *causal edges assumption* encodes causal relationships between variables by declaring that each parent node acts as a direct cause of its children [@Neal_2020, pp. 22]. @fig-ACflow illustrates how these assumptions influence the statistical and causal interpretations of graphs.

![ The flow of association and causation in graphs. Extracted and slightly modified from @Neal_2020 [pp. 31] ](images/png/ACflow.png){#fig-ACflow fig-align="center" width=80%}

>>>>

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines causal and Bayesian inference methods. In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear throughout this document clarifies these concepts. In this example, researchers aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?".

Following current CJ practices, researchers rely on estimates from the BTL model--or its transformations--to approximate this estimand. However, as discussed in @sec-theory-issues, Thurstone's Case V and the BTL model exhibit several statistical and measurement limitations. These limitations hinder the model's ability to identify various estimands relevant to CJ inquiries, including the one described in the example.

@@

<!-- 6. What the section will do? -->
Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to causal inference. Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as in Equation ([-@eq-mat11]). Here, each element of $I$ represents a unique individual $i$ or $h$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $i, h$ with stimuli $a, b$. As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a (repeated) judgment index $k$. Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.
$$
I = \begin{bmatrix}
1 \\
\vdots \\
i \\
\vdots \\
h \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
i & a \\
\vdots & \vdots \\
h & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A}
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K}
\end{bmatrix}
$$ {#eq-mat11}

Additionally, we construct the matrix $R$ to map each row of the $IA$ matrix with a corresponding row from the $JK$ matrix. This matrix has $n$ rows and $6$ columns, where $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of stimuli generated by each pair of individuals in the population. Thus, we define the matrix as in Equation ([-@eq-mat12]). 
$$
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
i & a & h & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
\end{bmatrix}
$$ {#eq-mat12}

It is easier to visualize the structure of the previously defined vectors and matrices by considering an example. Assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the vectors and matrices described in Equations ([-@eq-mat11]) and ([-@eq-mat12]) take the form as in Equation ([-@eq-mat13]).

$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
$$ {#eq-mat13}

@@

We can illustrate the structure of these vectors more clearly with an example. Suppose researchers exclude the second student, the second text from each student, and the third judge from the setup shown in Equation ([-@eq-mat13]). Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the resulting vectors would have the following structure:

$$
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
1 \\
1 \\
1 \\
1 \\
0 \\
0 \\
0 
\end{bmatrix}
$$ {#eq-mat24}

@@

Finally, we define the *sample mechanism* $S$ in Equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. Each element $s_{(i,a,h,b,j,k)}$ is a binary value indicating the presence or absence of data rows in the matrix $R$ resulting from the sample mechanism, as in Equation ([-@eq-mat26]). Thus, the vector contains $n$ elements, matching the number of rows in $R$, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.

@@

$$
S = \begin{bmatrix}
s_{(1,1,1,2,1,1)} \\
\vdots \\
s_{(1,1,1,2,1,n_{K})} \\
\vdots \\
s_{(i,a,h,b,j,k)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
$$ {#eq-mat25}

@@

<!-- 4. General summary of what we have done. -->
Thus, to address these limitations of Thurstone's Case V and the BTL model, this study extends Thurstone's general form using a systematic, integrated approach that combines causal and Bayesian inference methods. The approach begins by formulating a conceptual model represented by a Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This model integrates Thurstone's core theoretical principles, such as the discriminal processes of stimuli, alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms, thereby disentangling the causal processes in the CJ system.

@@

<!-- 5. Are there any additional advantages? -->
Beyond these potential benefits, the approach offers two additional advantages. First, it clarifies the roles and interactions of all actors and processes involved in CJ experiments. Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data. Together, these advantages establish a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022], offering new insights into existing research and opening promising avenues for future inquiry. Five current and new research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences. The following sections outline these avenues and explain how our model provides a rigorous research framework to investigate them.

@@

@@

<!-- 1. Challenges for the framework: background knowledge and research context -->
Despite its promise, this study presents several challenges that researchers must navigate--challenges common to many quantitative studies. They fall into two main categories. The first concerns the foundational knowledge researchers need to apply the proposed framework effectively. The second involves the increased attention researchers must now devote to their specific research context.

### Foundational knowledge {#sec-discussion_challenges1}

Researchers must develop foundational knowledge in two critical areas: causal and Bayesian inference methods. One example that illustrates the importance of mastering causal inference is our common assumption that predictor variables are "relevant" to the research context--interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference. As a result, to apply the framework rigorously, researchers must engage deeply with these and other complex ideas, such as how SCMs encode functional and probabilistic information. To assist this effort, we have provided key references throughout the study to guide motivated readers toward a deeper understanding of these foundational concepts[^9].

Additionally, researchers must develop the expertise to translate context-specific conceptual models into bespoke statistical models using Bayesian inference methods. While Bayesian approaches introduce specific technical demands--such as familiarity with probabilistic programming languages (PPLs) and concepts like probability distributions and convergence--we argue they offer a more accessible path for many applied researchers. This is because Bayesian methods significantly reduce the need for specialized knowledge in areas like optimization theory and techniques, which are often necessary in frequentist approaches.

### Attention to the research context {#sec-discussion_challenges2}

When applying our framework to a specific CJ research context, researchers face two additional challenges. First, they must verify that the conceptual model provides a *faithful* [^10] representation of the CJ system under study. Second, they must evaluate whether the model's statistical translation can accurately estimate the intended estimands (i.e., parameters) from empirical data. Notably, addressing these challenges requires careful attention because causal inference relies on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025]. Consequently, the framework does not deliver automatic answers; instead, it encourages researchers to formulate precise questions and explicitly state their assumptions to build a generalizable understanding of the phenomenon under study, in this case, the CJ system [@Rohrer_et_al_2022; @Deffner_et_al_2022]. This theory-driven foundation then supports the empirical testing of causal assumptions [@Deffner_et_al_2022].

[^10]: Avoid confusing this term with the *faithfulness assumption* described in the causal inference literature. For further details, see footnote 1.

<!-- Causal inference generally requires expert knowledge and untestable assumptions about the causal network linking treatment, outcome, and other variables. ...
Because the scenarios we considered were so simple, there was really no need to make the causal network explicit. As we start to turn our attention towards more complex situations, however, it will become crucial to be explicit about what we know and what we assume about the variables relevant to our particular causal inference problem. [@Hernan_et_al_2025, pp. 73] -->

<!-- Causal DAGs are of no practical use unless we make an assumption linking the causal structure represented by the DAG to the data obtained in a study. This assumption, referred to as the causal Markov assumption, states that, conditional on its direct causes, a variable Vj is independent of any variable for which it is not a cause. That is, conditional on its parents, Vj is independent of its non-descendants; hence, a causal DAG is Markov with respect to the DAG G. [@Hernan_et_al_2025, pp. 74] -->

<!-- Causal diagrams are a simple way to encode our subject-matter knowledge, and our assumptions, about the qualitative causal structure of a problem. But, as described in the next sections, causal diagrams also encode information about potential associations between the variables in the causal network. It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool. [@Hernan_et_al_2025, pp. 75] -->

@@

Thus, regarding the *faithful* representation of the CJ system, we advise researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets. This consideration means that researchers must face the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification. This careful adaptation is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019]. Nevertheless, as @Kohler_et_al_2019 points out, drawing conclusions about the real world from observed data always requires assumptions, regardless of whether the data are observational or experimental. In this context, our approach provides a transparent framework for researchers to clearly articulate these assumptions and design CJ experiments that support reliable and valid trait measurement.

@@

@@

@@

Finally, regarding the estimation capabilities of the statistical model, researchers must tackle the challenge of performing identification analysis. As outlined in @sec-theoretical, *identification analysis* determines whether a statistical model can accurately compute a given estimand (e.g., a parameter) based solely on its (causal) assumptions, independent of random variability [@Schuessler_et_al_2023]. Identification is crucial because it is a necessary condition for consistency. *Consistency* is the property of an estimator (or model) whose estimates converge to the "true" value of an estimand as data size approaches infinity [@Everitt_et_al_2010]. Logically, if researchers fail to identify a model, it cannot achieve consistency--even with infinite, error-free data. As a result, deriving meaningful insights from finite data becomes impossible [@Schuessler_et_al_2023].

Although performing identification analysis through formal derivations may appear to be a natural next step, we argue that the complexity of the CJ system renders such an approach impractical at the outset. Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs. Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research.

@@

<!-- 1. How the SCMS and DAGs encode probabilistic information? -->
Beyond their graphical capabilities, researchers can translate SCMs and DAGs into statistical models because these conceptual models encode functional and probabilistic information, which they can then replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016]. Notably, this encoding relies on two fundamental assumptions: the causal Markov and the faithfulness assumption. The *causal Markov assumption* links the graph structure to the joint probability distribution of variables by asserting that any variable in a graph is independent of all its non-descendants conditional on its parents [@Neal_2020, pp. 20; @Hernan_et_al_2025, pp. 80]. The *faithfulness assumption*, in turn, states that the observed statistical (in)dependencies in the data reflect the underlying causal structure among variables--except in rare degenerate cases [@Neal_2020, pp. 100; @Hernan_et_al_2025, pp. 81].

<!-- Causal DAGs are of no practical use unless we make an assumption linking the causal structure represented by the DAG to the data obtained in a study. This assumption, referred to as the causal Markov assumption, states that, conditional on its direct causes, a variable Vj is independent of any variable for which it is not a cause. That is, conditional on its parents, Vj is independent of its non-descendants; hence, a causal DAG is Markov with respect to the DAG G. [@Hernan_et_al_2025, pp. 74] -->

<!-- causal diagrams also encode information about potential associations between the variables in the causal network. It is precisely this simultaneous representation of association and causation that makes causal diagrams such an attractive tool. [@Hernan_et_al_2025, pp. 75] -->

<!-- This and the previous section review three structural reasons why two variables may be associated: one causes the other, they share common causes, or they share a common effect and the analysis is restricted to certain level of that common effect (or of its descendants). Along the way we introduced a number of graphical rules that can be applied to any causal diagram to determine whether two variables are (conditionally) independent. The arguments we used to support these graphical rules were heuristic and relied on our causal intuitions. These arguments, however, have been formalized and mathematically proven. See Fine Point 6.1 for a systematic summary of the graphical rules, and Fine Point 6.2 for an introduction to the concept of faithfulness. [@Hernan_et_al_2025, pp. 79] -->

<!-- (from probability to graph): The relationship between statistical independence and the purely graphical concept of d-separation relies on the causal Markov assumption (Technical Point 6.1): In a causal DAG, any variable is independent of its non-descendants conditional on its parents. Pearl (1988) proved the following fundamental theorem: The causal Markov assumption implies that, given any three disjoint sets A, B, C of variables, if A is d-separated from B conditional on C, then A is statistically independent of B given C. The assumption that the converse holds, i.e., that A is d-separated from B conditional on C if A is statistically independent of B given C, is a separate assumption—the faithfulness assumption described in Fine Point 6.2. [@Hernan_et_al_2025, pp. 80] -->
